---
title: "Practical Machine Learning assignment"
output: html_document
---


Introduction
------------

The goal of this project is to predict how well 6 different people performed barbell lifts utilizing data collected from activity monitoring devices. Each of the 6 people were asked to perform the barbell lifts correctly and in 5 different incorrect ways. Utilizing the activity monitor device data, a machine learning model is to be generated using a training set with class labels representing the 6 ways of performing the barbell lifts (supervised learning). Once the models are built, the generalization performance should be assessed, and then the training model is to be applied to a new set of testing data to make predictions. These predictions are later submitted for automated grading in a second component of the assignment

Data Analysis and Predictions
-----------------------------

The caret package is used here, which provides a streamlined interface into a variety of machine learning methods, making the entire analysis process much easier.

In addition, the Hmisc, randomForest, foreach, and doParallel libraries we loadad. This is needed to utilize multiple cores which will be used during the modeling building cross validation procedure to speed-up the calculations

```{r, echo=TRUE}
library(Hmisc)
library(caret)
library(randomForest)
library(foreach)
library(doParallel)
set.seed(2048)
options(warn=-1)
```

The data was loaded for from the provided training and test data. Some values contained a "#DIV/0!" that I replaced with an NA value.

```{r}
training_data <- read.csv("pml-training.csv", na.strings=c("#DIV/0!") )
evaluation_data <- read.csv("pml-testing.csv", na.strings=c("#DIV/0!") )
```

All columns 8 were converted to numeric

```{r, echo=FALSE}
for(i in c(8:ncol(training_data)-1)) {training_data[,i] = as.numeric(as.character(training_data[,i]))}

for(i in c(8:ncol(evaluation_data)-1)) {evaluation_data[,i] = as.numeric(as.character(evaluation_data[,i]))}

```

Because some columns were mostly blank, this did not contribute to the prediction. I chose a feature set that only included complete columns.

```{r}
feature_set <- colnames(training_data[colSums(is.na(training_data)) == 0])[-(1:7)]
model_data <- training_data[feature_set]
feature_set

```

We now have the model data built from our feature set.
```{r}
idx <- createDataPartition(y=model_data$classe, p=0.75, list=FALSE )
training <- model_data[idx,]
testing <- model_data[-idx,]

```

5 random forests with 150 trees each werebuilt. Parallel processing was used to build this model.

```{r}
registerDoParallel()
x <- training[-ncol(training)]
y <- training$classe

rf <- foreach(ntree=rep(150, 6), .combine=randomForest::combine, .packages='randomForest') %dopar% {
randomForest(x, y, ntree=ntree) 
}

```

Provide error reports for both training and test data.

```{r}
predictions1 <- predict(rf, newdata=training)
confusionMatrix(predictions1,training$classe)


predictions2 <- predict(rf, newdata=testing)
confusionMatrix(predictions2,testing$classe)

```



Prepare the submission. (using COURSERA provided code)
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}


x <- evaluation_data
x <- x[feature_set[feature_set!='classe']]
answers <- predict(rf, newdata=x)

answers

pml_write_files(answers)

```

